\textbf{\sc Towards Deliberative Control of Autonomous Robots}

### 2012-02-07 europa.tex
updates

\section{EUROPA}
\label{sec:europa}

EUROPA is a general purpose AI planning toolkit developed at NASA Ames.  Automated Planning can be reformulated as a constraint satisfaction problem (TODO: ref). EUROPA is an extension of that idea, however, unlike other approaches where the entire planning problem is translated into a CSP and then solved using traditional CSP methods, EUROPA uses a constraint reasoning engine as a fundamental building block, and enriches it with other architectural components that allow a more direct mapping for scheduling and planning problems, which in turn allows easier development of efficient solving methods.

Let's look at some of the problems that can be tackled using EUROPA.

Constraint Satisfaction: 
In the N-Queens problem, chess queens must be placed on an NxN chessboard in a way that there are no queens attacking each other.

Below is an example of a random positioning of queens on a NxN chessboard. Queens in violation of the non-attack constraint are highlighted in red.

\begin{figure}
\centering
\includegraphics[scale=0.35]{figs/Example-NQueens0.jpg}
\caption{\small N-Queens problem}
\label{fig:nqueens-1}
\vskip+0.1cm
\end{figure}

If we define NxN variables Qrc, r E [1,N], c E[1,N] , 
Qrc = 1 if Cell r,c in he chessboard is occupied by a Queen, 0 otherwise.

then we have to satisfy the following constraints:
Sum(Qrc) = 1 ForAll r (Only one Queen per row)
Sum(Qrc) = 1 ForAll c (Only one Queen per column)

Sum(Qr+i,c+i) = 1
Sum(Qr-1,c+i) = 1 (only on Queen on each diagonal)

We can solve the problem by finding assignments for all variables Qrc that satisfy the constraints above.

Below is a solution found by EUROPA using that formulation and a specialized search procedure.

\begin{figure}
\centering
\includegraphics[scale=0.35]{figs/Example-NQueens1.jpg}
\caption{\small N-Queens solution}
\label{fig:nqueens-2}
\vskip+0.1cm
\end{figure}

Scheduling:
In the Resource Constrained Project Scheduling Problem (RCPSP), a project consisting of a set of activities must be scheduled in a way that satisfies minimum and/or maximum temporal separation constraints. The activity schedule must also respect fixed limits on the availability of resources required to perform each activity. In addition to satisfying temporal and resource constraints, it is common for the user to want to minimize makespan so that the entire project is finished as early as possible.

Below is an example of a a solution provided by EUROPA for an RCPSP instance with 10 activities, 5 resources and 30 temporal constraints.

\begin{figure}
\centering
\includegraphics[scale=0.35]{figs/Example-UBO0.jpg}
\caption{\small RCPSP problem}
\label{fig:rcpsp-1}
\vskip+0.1cm
\end{figure}

Planning:
In the Shopping Agent Problem (TODO: ref to Norvig) an agent needs to purchase a set of products that are available at specific locations (supermarket, hardware store, etc), the agent  is subject to temporal (must complete tasks by specific deadlines) and resource (fuel, carrying capacity, etc) constraints. The agent must figure out what actions need to be performed to find and acquire the required items, as well as when to perform each of those actions.

Below is a solution produced by EUROPA for a simple problem instance where the shopping agent needs to buy Bananas, Milk and a Drill.

\begin{figure}
\centering
\includegraphics[scale=0.35]{figs/Example-Shopping0.jpg}
\caption{\small Shopping Agent problem}
\label{fig:shopping-1}
\vskip+0.1cm
\end{figure}

As we can see, a planning problem (what actions must be performed to achieve a goal) may embed a scheduling problem (in what order should those actions be performed), and both planning and scheduling embed Constraint Satisfaction problems since temporal, resource and many other kinds of constraints must be respected in most real life problems.

These relationships between Planning, Scheduling and Constraint Satisfaction have been examined in detail in the literature (TODOL ref Frank et al on Planning vs. Scheduling), and they lead to EUROPA's use of constraint reasoning as its innermost building block. Next we will examine the main ideas from Constraint Satisfaction Programming, and from Constraint Based Planning and Scheduling that constitute EUROPA's theoretical underpinnings, then we dive into how EUROPA implements those ideas.

\subsection{Constraint Programming}
\label{sec:europa:cp}

Constraint Satisfaction Programming, also know as Constraint Programming (CP) is a discipline that provides a generic framework for representing, solving and making logical inference on constraints. A complete treatment of this discipline is given in [MAR98] and [APT03] and very concise introductions are provided in [BAR99] and [LUS01].

	A constraint programming problem consists of a set of variables V={x1,..,xN}, where each variable takes values from a domain d1,..,dN. Domains are most often discrete and finite, but there are numerous CP implementations associated with continuous domains. Given a defined conjunctive set of constraints on the variables: $C=\{c1(x1,..,xN), ..., cK(x1,..,xN)\}$, the objective is to find one or more value assignments to V where all constraints are satisfied.

To solve a problem, CP uses logical inference to perform 2 operations:

\begin{enumerate}
	\item Bounds propagation: To infer upper and lower variable bounds. For example, from the constraints x1 + x2 <= 2, xi>=0, we can infer [0,2] bounds for x1 and x2

	\item Domain reduction: To infer a valid set of values for a variable. For example, for discrete variables, the constraints allDifferent(x1,x2,x3) and xi>=0, xi <=4. if x1 =1 and x2=3 we can infer that the valid domain for x3 is reduced to {2,4}

@@

CP is normally implemented as part of a programming language and constraints are normally represented as objects [PUG95]. Any constraint can be introduced and, as long as the bounds propagation and domain reduction protocols specified by the host CP system are enforced, it will be indistinguishable from any other ÒprimitiveÓ CP constraint, such as <= or >=.

Solving a CSP is NP-Hard in theory, but often very efficient in practice using a number of algorithms and techniques that are well understood.

TODO: talk about arc consistency algorithms, computational complexity and solvers.

\subsection{Constraint-Based Attribute and Interval Planning}
\label{sec:europa:cp}

There has been 
State Variable representation

CSP encodings

Constraint Based and interval planning.

\subsection{EUROPA's Architecture}
\label{sec:europa:arch}

The following figure displays the main architectural components in EUROPA and their relationships:

\begin{figure}
\centering
\includegraphics[scale=0.35]{figs/EUROPA-Architecture.jpg}
\caption{\small EUROPA's Architecture}
\label{fig:europa-architecture}
\vskip+0.1cm
\end{figure}

	\item The Constraint Reasoning Engine (CRE): manages variables, the domains from which they can take values, and constraints that define relationships among variables. It also provides an efficient arc consistency mechanism (TODO: ref to AC-3). The CRE is designed so that specialized reasoning algorithms for specific constraints can be easily and efficiently plugged in.

	\item The Plan Database (PDB): manages object types and objects, which are a mechanism to group variables to more naturally model the real world in much the same way that object oriented programming does. Also manages tokens, which are a mechanism to group variables to represent temporally scoped state. Objects and tokens can be used to model planning domains in a much more natural and extensible way that a pure CSP approach can (TODO: provide snippet?)

	TODO: talk about the token network, move stuff from the Concepts section if necessary

	\item The Rules Engine (RE): manages causal dependencies in a planning model (TODO: explain)

	\item The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)

	\item API Layer: makes all the services available so that EUROPA can be used to build applications.

	\item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
		\begin{enumerate}
			\item Temporal Network
			\item Resources
			\item NDDL
			\item ANML (TODO:maybe it's better to leave out?)
		\end{enumerate}

EUROPA can ingest descriptions of models, plans and goals. This means that a single planner can be applied to different systems and problems, simply by providing different models and goals. For such planners the expressiveness of the languages for models, goals and plans is of great importance. EUROPA provides a high level language called NDDL (New Domain Description Language) that allows the user to specify the different components of a problem domain in a concise way. The main characteristics that make NDDL a powerful tool for describing problem domains are:

\begin{enumerate}
	\item It's Object Oriented: NDDL supports classes and inheritance in similar fashion to popular Object-Oriented languages like Java and C++. It also supports polymorphism for some of the planning components. Using object classes and instances is a time-tested approach to naturally describe problem domains (TODO: OO ref?).
	\item It offers declarative constructs to define constraints and causality dependencies between actions and effects.
	\item It offers procedural constructs to populate EUROPA's plan database with specific problem instances expressed in terms of objects, variables, constraints, facts and goals.
\end{enumerate}

@@

  starts(effect Own purchase);
  eq(purchase.product,product);
}

TODO: explain variable, object, predicate and action types, including constraints that can be part of them

@@

\begin{enumerate}
	\item Variable Types:
	\item Object Types:
	\item Predicate Types:
	\item Action Types:
\end{enumerate}

Now let's look at how a specific problem instance would be specified in NDDL:

@@

TODO: explain how inference is used to detect perform bounds propagation, domain reduction and to detect constraint violations.
- talk about AC3 implementation and extension mechanisms
- talk about europa's constraint library

The Temporal Network:
The notion of time is central to temporal planning. EUROPA uses variables to explicitly represent timepoints for plan activities and states. Constraints among timepoints provide a natural way to express domain axioms. For example, in order to state that activity A must occur before activity B we can say that the end timepoint of A is <= the start timepoint of B. Dechter et al. (Dechter 1991) proposed that constraints among timepoints can be grouped together to form a Simple Temporal Network (STN). Such a network can be transformed into a Distance Graph (DG) where the outward arc from a node represents the maximum distance from the source node to the target node. The diagram below illustrates a simple STN with just 2 variables and a single constraint. It also shows the resulting DG.

TODO: add figure

Dechter et al. (Dechter 1991) also showed that shortest path algorithms could be used to propagate values in the network and discover a negative cycle. A negative cycle is a path from a node to itself that has a path length less than 0. If such a cycle exists, the network is inconsistent. It was further shown that a single-source shortest path algorithm was sufficient to detect a negative cycle and provide sufficient propagation to yield a backtrack-free search. Thus we have an efficient and complete algorithm for propagating an STN. These results build on the already established notion of a CSP and are naturally incorporated into the general representation and propagation scheme used in EUROPA.

@@

EUROPA provides a built-in solver that performs Plan Space Planning (TODO: ref), in this approach, the initial state is considered a partial plan that needs to be refined toward a solution plan that achieves the goals. The operations to refine the partial plan PP at any time are:
\begin{enumerate}
	\item Find the flaws of PP, that is the conditions that prevent it from being a solution plan (flaw types are explained below).
	\item Select one such flaw
	\item Select a resolver for the flaw
	\item Refine PP by applying the resolver
	\item If an inconsistency is found, try another resolver.
	\item If all possible resolvers for a particular flaw fail, return failure, otherwise continue until resolving all the flaws.
\end{enumerate}

This Plan Space Planning algorithm is translated into EUROPA's representation as follows:

The initial partial plan is the state of EUROPA's Plan Database after the initial state has been instantiated, this results in a set of variable, object and token instances. Inference then takes place to detect flaws in the partial plan, there are 3 kinds of flaws that can be detected:

\begin{enumerate}
	\item Unbound Variable: a variable in the partial plan whose domain is not a singleton. Unbound Variables are resolved by specification of a value from the domain of the variable.

	\item Open Condition: an open condition is an inactive token. Inactive tokens can be generated by explicitly posted goals, or when a token is activated, rules may fire that create inactive slave tokens (TODO: point to more detailed decription above in Rules Engine section).  Open Conditions can be resolved in three ways: 
\begin{enumerate}
		\item Merging: A free activity is merged with a matching activity already in the plan. An activity a is said to match an activity a? if a and a? unify and the temporal constraints involving a are satisÞed by a?. Thus,  a and a? can be considered the same activity and we do not need to introduce a in the plan. Consequently, the compatibilities associated with a are not Þred, because they have been already triggered when a? was introduced in the plan. 
		\item Activation: We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciÞc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities. This results in both an ordering ßaw, corresponding to the just activated activity, and a number of open condition ßaws, corresponding to the new subgoal activities. 
		\item Rejection: Some goals may be optional, if that is the case for a particular order condition, its corresponding flaw can be resolved by discarding the inactive token.
\end{enumerate}

@@

The Plan Space Planning algorithm described above can be implemented in many different ways depending on the approach chosen for flaw and resolver selection and for backtracking. EUROPA's built-in solver implements a chronological backtracking algorithm \ref{alg:europa:solve}
 % summarized in the figure below.

@@

% \begin{verbatim}

% 01.  bool solve(PartialPlan plan)
% 02.   {
% 03.      if (isInconsistent(plan))  return false;
% 04.
% 05.      Flaw flaw = chooseFlaw(plan);
% 06.      if (flaw == NULL) return true;
% 07.
% 08.      DecisionPoint decision = makeDecisionPoint(flaw, plan);
% 09.
% 10.      while(decision.hasNext()) {
% 11.          PartialPlan pp = decision.executeNext();
% 12.
% 13.          if (solve(pp)) return true;
% 14.          else d.undo();
% 25.      }
% 16.
% 17.      return false;
% 18.  }

% \end{verbatim}

@@

TODO: expand a little more on how heuristics can be implemented by applying ordering/filtering for chooseFlaw and makeDecisionPoint
examples:
- deal with earliest flaws first
- deal with planning (open condition flaws) before scheduling (threats) ones

It should be emphasized that while this algorithm is commonly employed, it is only one of many that could be implemented. TODO: give examples of local search solvers for scheduling and CSP?

>>>>

\subsection{\eu Architecture}
\label{sec:europa:arch}

\begin{figure}
\centering
\includegraphics[scale=0.5]{figs/EUROPA-Architecture.jpg}
\caption{\small The \eu Architecture}
\label{fig:europa-architecture}
\vskip+0.1cm
\end{figure}

Fig. \ref{fig:europa-architecture} shows the main architectural
components in \eu and their relationships.

The Constraint Reasoning Engine (CRE) manages variables, the domains
from which they can take values, and constraints that define
relationships among variables. It also provides an efficient arc
consistency mechanism \cite{mackworth77}. The CRE is designed so that
specialized reasoning algorithms for specific constraints can be easily
and efficiently plugged in. The Plan Database (PDB) manages object types
and objects, which are a mechanism to group variables to more naturally
model the real world in much the same way that object oriented
programming does. Also manages tokens, which are a mechanism to group
variables to represent temporally scoped state. Objects and tokens can
be used to model planning domains in a much more natural and extensible
way that a pure CSP approach can. The Rules Engine (RE) manages token
relationships in a planning model. The language modules (NDDL, ANML, and
other higher level modeling languages) are used to create domain models
and problem instances \comment{point advantages over pure programming
  API approaches, provide snippets}. Finally the API Layer (Fig.
\ref{fig:europa-api}) makes all the services available so that \eu can
be used to build applications.

\comment{Is Fig. \ref{fig:europa-api} necessary?}

\begin{figure}
\centering
\includegraphics[scale=0.35]{figs/EUROPA-API.jpg}
\caption{\small EUROPA's Client API}
\label{fig:europa-api}
\vskip+0.1cm
\end{figure}

	% \item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
	% 	\begin{enumerate}
	% 		\item Temporal Network
	% 		\item Resources
	% 		\item NDDL
	% 		\item ANML (TODO:maybe it's better to leave out?)
	% 	\end{enumerate}

\eu can ingest descriptions of models, plans and goals and thereby can
be applied to different systems and problems, simply by providing
different models and goals. Consequently the expressiveness of the
languages for models, goals and plans is of great importance. For \eu
NDDL (New Domain Description Language [pronounced ``noodle'']
\comment{cite}) allows the user to specify the different components of a
problem domain in a concise way. The main characteristics that make NDDL
a powerful tool for describing problem domains are:

\item Object Oriented: NDDL supports classes and
  inheritance in similar fashion to popular Object-Oriented languages
  like Java and C++. It also supports polymorphism for some of the
  planning components. Using object classes and instances is a
  time-tested approach to naturally describe problem domains \comment{OO
    citation}.

\item It offers declarative constructs to define constraints
  and causality dependencies between actions and effects.

\item It offers procedural constructs to populate \eus plan database
  with specific problem instances expressed in terms of objects,
  variables, constraints, facts and goals.

@@

@@

  starts(effect Own purchase);
  eq(purchase.product,product);
}
\end{verbatim}

@@

\comment{explain variable, object, predicate and action types, including constraints that can be part of them}
\begin{enumerate}
	\item Variable Types: This model uses integer (activity duration, start and end times, etc) and string (Location and Product names, etc)
	\item Object Types:
	\item Predicate Types:
	\item Action Types:
\end{enumerate}

 shows how a specific problem instance would be
specified in NDDL.

@@

\comment{explain how inference is used to detect perform bounds
  propagation, domain reduction and to detect constraint violations.} 
- talk about AC3 implementation and extension mechanisms
- talk about europa's constraint library

The Temporal Network: The notion of time is central to temporal
planning. \eu uses variables to explicitly represent timepoints for plan
activities and states. Constraints among timepoints provide a natural
way to express domain axioms. For example, in order to state that
activity A must occur before activity B we can say that the end
timepoint of A is $leq$ the start timepoint of B.

\begin{figure}
\centering
\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
\caption{\small Temporal relations defined within the planner are
  based on \texttt{Allen Algebra} relations shown above.}
\label{fig:allen-algebra}
\vskip-0.3cm
\end{figure}

\cite{dechter91} proposed that constraints among timepoints can be
grouped together to form a Simple Temporal Network (STN). Such a network
can be transformed into a Distance Graph (DG) where the outward arc from
a node represents the maximum distance from the source node to the
target node. They also showed that shortest path algorithms could be
used to propagate values in the network and discover a negative cycle. A
negative cycle is a path from a node to itself that has a path length
less than 0. If such a cycle exists, the network is inconsistent. It was
further shown that a single-source shortest path algorithm was
sufficient to detect a negative cycle and provide sufficient propagation
to yield a backtrack-free search. Thus we have an efficient and complete
algorithm for propagating an STN. These results build on the already
established notion of a CSP and are naturally incorporated into the
general representation and propagation scheme used in \eu.
The diagram below illustrates a simple STN with just 2
variables and a single constraint. It also shows the resulting DG.

\comment{ add figure}

@@

\eu provides a built-in solver that performs Plan Space Planning
\cite{ghallab04} in this approach, the initial state is considered a
partial plan that needs to be refined toward a solution plan that
achieves the goals. The operations to refine the partial plan PP at any
time are:

\begin{enumerate}
	\item Find the flaws of PP, that is the conditions that prevent it from being a solution plan (flaw types are explained below).
	\item Select one such flaw
	\item Select a resolver for the flaw
	\item Refine PP by applying the resolver
	\item If an inconsistency is found, try another resolver.
	\item If all possible resolvers for a particular flaw fail, return failure, otherwise continue until resolving all the flaws.
\end{enumerate}

This Plan Space Planning algorithm is translated into \eus
representation as follows:

The initial partial plan is the state of \eus Plan Database after the
initial state has been instantiated, this results in a set of variable,
object and token instances. Inference then takes place to detect flaws
in the partial plan, there are 3 kinds of flaws that can be detected:

\item Unbound Variable: a variable in the partial plan whose domain is
  not a singleton. Unbound Variables are resolved by specification of a
  value from the domain of the variable.

\item Open Condition: an open condition is an inactive token. Inactive
  tokens can be generated by explicitly posted goals, or when a token is
  activated, rules may fire that create inactive slave tokens
  \comment{point to more detailed decription above in Rules Engine
    section}.
  Open
  Conditions can be resolved in three ways:

  \item Merging: A free activity is merged with a matching activity
    already in the plan. An activity a is said to match an activity a?
    if a and a? unify and the temporal constraints involving a are
    satisÞed by a?. Thus, a and a? can be considered the same activity
    and we do not need to introduce a in the plan. Consequently, the
    compatibilities associated with a are not Þred, because they have
    been already triggered when a? was introduced in the plan.

  \item Activation: We introduce a new activity a in the current plan
    associating it with the proper timeline, but without choosing a
    speciÞc time slot for it.The compatibilities associated with a are
    applied and the subgoal activities resulting from those
    compatibilities are introduced as free activities. This results in
    both an ordering flaw, corresponding to the just activated activity,
    and a number of open condition flaws, corresponding to the new
    subgoal activities.

  \item Rejection: Some goals may be optional, if that is the case for a
    particular order condition, its corresponding flaw can be resolved
    by discarding the inactive token.

@@

The Plan Space Planning algorithm described above can be implemented in
many different ways depending on the approach chosen for flaw and
resolver selection and for backtracking. \eus built-in solver implements
a chronological backtracking algorithm that is summarized in Algorithm
\ref{alg:europa:solve}.

@@

@@

\comment{expand a little more on how heuristics can be implemented by
  applying ordering/filtering for chooseFlaw and makeDecisionPoint}
examples:
- deal with earliest flaws first
- deal with planning (open condition flaws) before scheduling (threats) ones

It should be emphasized that while this algorithm is commonly employed,
it is only one of many that could be implemented. \comment{give examples
  of local search solvers for scheduling and CSP?}
